# Welcome {-}

```{r "intro"}
#| echo: false
.gh_url <- file.path('https://github.com', rmarkdown::metadata[['github-repo']])
pkglines <- read.dcf("DESCRIPTION")
pkg <- pkglines[, "Package"][[1]]
version <- pkglines[, "Version"][[1]]
authors <- eval(parse(text = pkglines[, "Authors@R"]))
license <- pkglines[, "License"][[1]]
intro <- glue::glue(
    "**Package:** {pkg}<br/>\n", 
    "**Authors:** {paste(format(authors, include = c('given', 'family', 'role')), collapse = ', ')}<br/>\n", 
    "**Compiled:** {as.character(Sys.Date())}<br/>\n", 
    "**Package version:** {version}<br/>\n", 
    "**R version:** <b style='color:red;'>{R.version.string}</b><br/>\n", 
    "**BioC version:** <b style='color:red;'>{BiocManager::version()}</b><br/>\n", 
    "**License:** {license}<br/>"
)
```

`r intro`

This is the landing page for the **"Epigenomics Data Analysis"** 
workshop, ed. 2025.  

## What {-}

This course will introduce biologists and bioinformaticians to the field of regulatory epigenomics. 
We will cover a range of software and analysis workflows for processing and quantitative analysis of 
ChIP-seq, MNase and ATAC-seq data, RNA-seq, and Hi-C data. We will start by introducing general 
concepts about epigenomics. From there, we will then continue to describe the main analysis 
steps to go from raw sequencing data to processed and usable data. We will present classical 
analysis workflows, their output and the possible paths to investigate downstream of this. 
Towards the end of the workshop, we will focus on multi-omics data integration. 
Throughout the workshop, real-world datasets will be analysed using bash tools and R/Bioconductor packages.

## Target audience {-}

This course is primarily targeting researchers who are relatively new to the field 
of bioinformatics, with practical experience in the experimental side of epigenomics. 
Attendees should have a background in biology as well as be familiar with genomic data 
and common file formats from NGS sequencing experiments (fastq, BAM, BED).  

Practical exercises will use command-line Linux and R code and will be presented as notebooks to ensure reproducible coding.
Attendees are expected to know the basics of the R programming language and have prior expe- rience with the Unix command line.

## Why {-}

By the end of the course, participants will be able to:

- Understand the principles and experimental designs of key genomic assays.
- Perform preprocessing, alignment, and quality control for RNA-seq, ATAC-seq, ChIP-seq, and Hi-C data.
- Conduct downstream analyses such as differential expression, peak calling, motif analysis, and interaction visualization.
- Integrate multiple genomic data types to derive biological insights.

## Instructors {-}

[Jacques Serizay](mailto:jacques.serizay@pasteur.fr)

<!-- # Docker image {-}

A `Docker` image built from this repository is available here: 

👉 [ghcr.io/js2264/ngs-physalia](https://ghcr.io/js2264/ngs-physalia) 🐳

::: {.callout-tip icon='true'}

## Get started now 🎉

You can get access to all the packages used in this book in < 1 minute, 
using this command in a terminal: 

```{sh "docker", filename="bash"}
#| eval: false
docker run -it ghcr.io/js2264/ngs-physalia:devel R
```

::: -->

<!-- # RStudio Server {-}

An RStudio Server instance can be initiated from the `Docker` image as follows: 

```{sh "rstudio", filename="bash"}
#| eval: false
docker run \
    --volume <local_folder>:<destination_folder> \
    -e PASSWORD=OHCA \
    -p 8787:8787 \
    ghcr.io/js2264/ngs-physalia:devel
```

The initiated RStudio Server instance will be available at 
[https://localhost:8787](https://localhost:8787). -->

# Session info {-}

::: {.callout-note collapse="true"}

## Click to expand 👇

```{r "session info"}
#| cache: false
sessioninfo::session_info(
    installed.packages()[,"Package"], 
    include_base = TRUE
)
```

:::
