---
engine: knitr
execute:
    eval: false
---

# Lab 2: Quantification and differential gene expression by RNA-seq

::: {.callout-note icon='true'}
## Aims

- Estimate transcript abundance from RNA-seq samples
- Import count data in R
- Perform differential expression analysis with `DESeq2`
:::

::: {.callout-tip icon='true'}
## Datasets

RNA-seq data was published in [Nuño-Cabanes et al., Scientific Data 2020](https://www.nature.com/articles/s41597-020-0412-z)

- Control RNA-seq @ 30 C: 

    - `SRR9929263`
    - `SRR9929264`
    - `SRR9929273`
    - `SRR9929282`

- Heat shock RNA-seq @ 39 C, 20 min:

    - `SRR9929271`
    - `SRR9929265`
    - `SRR9929280`
    - `SRR9929274`
:::


## Count mapped fragments over a set of gene annotations 

- Check the files in the `Share/RNA` directory.

::: {.callout-answer .icon .callout-note collapse=true}
```{sh, filename = 'bash'}
ls -l Share/RNA/RNAseq*bam
```
:::

- Locate the gene annotation file provided for the yeast genome. 
- How many annotations are there in the file?
- What type of annotations are there? 
- How can you have more start codons than stop codons?

::: {.callout-answer .icon .callout-note collapse=true}
```{sh, filename = 'bash'}
wc -l Share/genome/R64-1-1.gtf 
head Share/genome/R64-1-1.gtf
cut -f3 Share/genome/R64-1-1.gtf | sort | uniq -c
```
:::

- Check `featureCounts` help to understand the different options available.
- Which level of annotations should you use to count the number of mapped fragments over?
- What are the `featureCounts` options specific to paired-end reads? And to stranded RNA-seq?
- Use `featureCounts` to count the number of mapped fragments over the set of gene annotations.

::: {.callout-answer .icon .callout-note collapse=true}
```{sh, filename = 'bash'}
featureCounts --help
featureCounts \
    -a Share/genome/R64-1-1.gtf \
    -o RNAseq_counts.txt \
    -t transcript \
    -g gene_id \
    -s 2 \
    -p \
    --countReadPairs \
    -P -B -D 10000 -C \
    -T 2 \
    Share/RNA/RNAseq*R64-1-1.bam

cat RNAseq_counts.txt.summary
```
:::

## Import counts into R

- Create a tibble containing the following information for the four RNA-seq samples: 

    - Path to local `bam` file
    - The biological sample it corresponds to (e.g. `WT`, `HS`)
    - The biological replicate it corresponds to (e.g. `rep1`, `rep2`, ...)

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(tidyverse)
samples <- tibble(
    bam = list.files('Share/RNA', pattern = 'RNAseq.*_R64-1-1.bam$', full.names = TRUE), 
    sample = basename(bam), 
) |> separate(sample, sep = '_', into = c(NA, 'condition', 'rep', NA)) |> 
    mutate(sample = paste(condition, rep, sep = '_'))
```
:::

- Import gene annotations for yeast. 
- Filter it out to only keep transcripts

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(rtracklayer)
genes <- import('Share/genome/R64-1-1.gtf')
genes <- genes[genes$type == 'transcript']
names(genes) <- genes$gene_id
```
:::

- Import the counts generated by `featureCounts` into R.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
cnts <- read_tsv('RNAseq_counts.txt', col_names = TRUE, comment = "#")
cnts <- cnts |> 
    select(-c(2:6)) |> 
    column_to_rownames('Geneid') |> 
    as.matrix() 
colnames(cnts) <- samples$sample
```
:::

- Wrap the counts, the sample metadata and the gene annotations into a `SummarizedExperiment` object.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(SummarizedExperiment)
se <- SummarizedExperiment(
    assays = list(counts = cnts), 
    rowRanges = genes, 
    colData = samples
)
```
:::

## Differential expression analysis

- Read `DESeq` documentation (`?DESeq`). What type of object does it require? How can you create one? 
- Create a `DESeqDataSet` object from the RNA counts. Choose your `design` formula appropriately.
- Run `DESeq` workflow 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(DESeq2)
dds <- DESeqDataSet(se, design = ~ condition)
dds <- DESeq(dds)
```
:::

- Check the documentation for the `results` function from `DESeq2` package.
- Extract the results of differential expression analysis, for the appropriate `contrast`. 
- Recover genes over-expressed in heat-shock vs. control growth (fold-change >= 2, p-value <= 0.01)

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
res <- results(dds, contrast = c('condition', 'HS', 'WT'))
upreg_genes <- res |> 
    as_tibble(rownames = 'gene') |> 
    filter(log2FoldChange > 1, padj <= 0.001)

upreg_genes |> pull(gene) |> writeLines()
```
:::
