---
engine: knitr
execute:
    eval: false
---

# Lab 4: ATAC-seq analysis

::: {.callout-note icon='true'}
## Aims

- Annotate ATAC-seq peaks
- Import ATAC-seq peaks and fragments in R
- Compare ATAC-seq and MNase-seq fragment sizes
:::

::: {.callout-tip icon='true'}
## Datasets

We will investigate two datasets from the Koszul lab, generated in 2024 and published [in Science](https://www.science.org/doi/10.1126/science.adm9466).

- [SRR31397086](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM8640705)
- [SRR31397088](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM8640707)
:::

## Annotate ATAC-seq peaks with yapc 

- Check documentation from the `yapc` package to see how to annotate peaks with `yapc`.
- Annotate ATAC peaks from the two provided `bw` files. 

#::: {.callout-note icon='true'}

`yapc` has been installed in a separate `micromamba` environment, because it is incompatible with other tools we will use in the workshop. 
We will use it to annotate ATAC-seq peaks.

:::

::: {.callout-answer .icon .callout-note collapse=true}
```{sh, filename = 'sh'}
micromamba run -n yapc_env yapc -h 
micromamba run -n yapc_env yapc atac wt Share/ATAC/ATAC_rep1.bw Share/ATAC/ATAC_rep2.bw
```
:::

- Inspect in `IGV` the peak sets generated with different p-value thresholds.
- Also check the `*d2smooth.bw` track. Can you understand how the peaks are identified? 

## Import ATAC-seq peaks in R

- Check documentation from the `rtacklayer` package to see how to import a `bed` file in `R`.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(tidyverse)
library(GenomicRanges)
peaks <- read_tsv('Share/ATAC/atac_0.05.bed', comment = "#", col_names = FALSE)
colnames(peaks) <- c("chr", "start", "end", "annot", "score", "strand", ".start", ".end", "NA")
peaks <- as(peaks, "GRanges")
peaks
```
:::

## Import ATAC-seq fragments in R

- Check documentation from the `Rsamtools` package to see how to create a connection to disk-stored `bam` files. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(Rsamtools)
atac_bam <- BamFileList(list(
    rep1 = BamFile('Share/ATAC/ATAC_rep1.bam'),
    rep2 = BamFile('Share/ATAC/ATAC_rep2.bam') 
))
atac_bam
```
:::

- Read `GenomicAlignments` package documentation to see how to import fragments from a `BamFile` connection. 
- Import fragments from **paired-end** reads, in **proper pairs**, **no duplicates** and **no secondary alignments**, with a **MAPQ >= 20**. The important `bam` column to recover is `isize` (insert size).

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(GenomicAlignments)
library(purrr)
param <- ScanBamParam(
    flag=scanBamFlag(
        isPaired = TRUE,
        isProperPair = TRUE,
        isDuplicate = FALSE,
        isSecondaryAlignment = FALSE
    ), 
    mapqFilter = 20, 
    what = c("isize")
)
atac_frags <- map(atac_bam, readGAlignmentPairs, param = param)
atac_frags
```
:::

## Check distribution of ATAC fragment sizes

- Coerce fragments to `GRanges` with the `as` function.
- Subset fragments to retain those overlapping imported ATAC-seq `peaks`.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
atac_frags <- map(atac_frags, as, "GRanges")
atac_frags <- map(atac_frags, subsetByOverlaps, peaks)
```
:::

- Plot the width distribution for filtered ATAC-seq fragments. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
df <- map_dfr(atac_frags, as_tibble) |> 
    group_by(width) |> 
    tally()
ggplot(df, aes(x = width, y = n)) + geom_line() + xlim(c(0, 600))
```
:::

- How can you interpret the resulting distribution?

## Compare ATAC-seq and MNase-seq fragment sizes

- Repeat the same procedure for MNase-seq fragments.

:::. {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
mnase_bam <- BamFile('Share/MNase/MNase_20_filtered_sorted.bam')
mnase_frags <- readGAlignmentPairs(mnase_bam, param = param)
mnase_frags <- as(mnase_frags, "GRanges")
df_mnase <- as_tibble(mnase_frags) |> 
    group_by(width) |> 
    tally()
df2 <- rbind(
    df |> mutate(type = 'ATAC-seq'),
    df_mnase |> mutate(type = 'MNase-seq')
)
ggplot(df2, aes(x = width, y = n, color = type)) + 
    geom_line() + 
    xlim(c(0, 600)) + 
    theme_bw()
```
:::

## Check distance from ATAC-seq to TSSs

- In Lab 2, we have imported genomic annotations in R. Use the same file to extract TSS positions.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(rtracklayer)
genes <- import('Share/genome/R64-1-1.gtf')
genes <- genes[genes$type == 'transcript']
mcols(genes) <- mcols(genes)[, c("gene_name", "gene_id", "gene_biotype")]
tss <- genes |> resize(width = 1, fix = 'start')
```
:::

- Check how many ATAC peaks overlap with TSSs.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
table(peaks %over% tss)
```
:::

- Check the distance from each ATAC to the nearest TSS.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
peaks <- plyranges::add_nearest_distance(peaks, tss)
df <- as_tibble(peaks)
quantile(df$distance, probs = c(0.1, 0.25, 0.5, 0.75, 0.9))
p <- ggplot(df, aes(x = distance)) + 
    geom_density(fill = 'lightblue', alpha = 0.5) +
    xlim(c(0, 1000)) +
    theme_bw()
```
:::
