---
engine: knitr
execute:
    eval: false
---

# Lab 5: Comparing sets of ChIP-seq peaks

::: {.callout-note icon='true'}
## Aims

- Import a dozen of ChIP-seq peak sets in R
- Check distribution of peaks compared to genomic features
- Check peak occurrence over tissue-specific regulatory elements
:::

::: {.callout-tip icon='true'}
## Datasets

- modENCODE/modERN TF ChIP-seq database, available [here](https://www.encodeproject.org/search/?type=Experiment&control_type!=*&status=released&replicates.library.biosample.donor.organism.scientific_name=Caenorhabditis+elegans&biosample_ontology.term_name=whole+organism&assay_title=TF+ChIP-seq&biosample_ontology.classification=whole%20organisms)
:::

## Download peaks from ENCODE database

Check out the [modENCODE/modERN TF ChIP-seq database](https://www.encodeproject.org/search/?type=Experiment&control_type!=*&status=released&replicates.library.biosample.donor.organism.scientific_name=Caenorhabditis+elegans&biosample_ontology.term_name=whole+organism&assay_title=TF+ChIP-seq&biosample_ontology.classification=whole%20organisms). 

Inspect the available datasets, by filtering e.g. using the following criteria: 

- Assay title: `TF ChIP-seq`
- Organism: `C. elegans`
- Genome assembly: `ce11`
- Project: `modENCODE`

Peak files for 12 ChIP-seq datasets have been already downloaded from this database. 

## Import all peaks in R

- In R, list all the `bed` files available for the peaks. 
- Import each file as a `GRanges` object. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(rtracklayer)
library(tidyverse)
bedfiles <- list.files('Share/peaks/modENCODE/', pattern = '*bed')
peaks <- map(bedfiles, ~ import(file.path('Share/peaks/modENCODE/', .x)))
names(peaks) <- str_replace(bedfiles, '.bed', '')
```
:::

## Define genomic features in `ce11`

Genomic features can be easily annotated from a set of gene features. 
`ChIPseeker` facilitates the annotation of ChIP-seq peaks using gene annotations 
directly provided in R by `TxDb` gene annotation packages, e.g. `TxDb.Celegans.UCSC.ce11.ensGene`

- Install the `TxDb.Celegans.UCSC.ce11.ensGene` package in R
- Install `ChIPseeker`

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'r'}
library(ChIPseeker)
library(TxDb.Celegans.UCSC.ce11.ensGene)
```
:::

## ChIP-seq peak overlaps with genomic features 

- Use `ChIPseeker` to annotate ChIP-seq peaks for a single ChIP-seq experiment. 
- Extract 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'r'}
ceh48 <- peaks[[1]]
annots_ceh48 <- annotatePeak(ceh48, TxDb = TxDb.Celegans.UCSC.ce11.ensGene)
getAnnoStat(annots_ceh48)
```
:::

- Iterate over all the peak sets to compile their annotations across *C. elegans* genome. To iterate over each set of peaks in the `peaks` list and return an aggregate `data.frame`, use the `imap_dfr`. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'r'}
annots <- imap_dfr(peaks, ~ annotatePeak(.x, TxDb = TxDb.Celegans.UCSC.ce11.ensGene) |> 
    getAnnoStat() |> 
    mutate(TF = .y)
)
annots
```
:::

- Generate a barplot to represent the % of peaks in each type of genomic features, for the 12 TFs. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'r'}
library(ggplot2)
p <- ggplot(annots, aes(x = TF, y = Frequency, fill = Feature)) + 
    geom_col() + 
    guides(x =  guide_axis(angle = 90))
```
:::

- Comment on the distribution of the 12 sets of peaks over genomic features. 

## Peaks occurrence over tissue-specific regulatory elements

- Import regulatory elements annotated in *C. elegans* in R, as seen in previous Lab. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(readxl)
download.file('https://genome.cshlp.org/content/suppl/2020/11/16/gr.265934.120.DC1/Supplemental_Table_S2.xlsx', 'WBcel235_REs.xlsx')
REs <- read_xlsx('WBcel235_REs.xlsx', skip = 2, col_names = TRUE)
REs <- makeGRangesFromDataFrame(
    REs, seqnames.field = 'chrom_ce11', 
    start.field = 'start_ce11', end.field = 'end_ce11',
    keep.extra.columns = TRUE
)
REs
```
:::

- Check for overlap of each regulatory element with `xnd-1` peaks. 
- Check the overlap of germline-specific regulatory element with `xnd-1` peaks. 
- Check the **enrichment** of germline-specific regulatory element over `xnd-1` peaks. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
table(REs %over% peaks[['xnd-1']])
table(REs %over% peaks[['xnd-1']], REs$Annotation)
table(REs %over% peaks[['xnd-1']], REs$Annotation == 'Germline') 
table(REs %over% peaks[['xnd-1']], REs$Annotation == 'Germline') |> fisher.test()
```
:::

- For each tissue (`Germline`, `Neurons`, `Muscle`, `Hypod.` and `Intest.`), check whether the tissue-specific REs are enriched in `xnd-1` peak.

::: {.callout-tip icon='true'}
- You can iterate over each `tissue` and return an aggregated `data.frame` using `map_dfr`. 
- For each iteration, you can transform the result of `fisher.test()` into a `tibble` with the `glance` function from `broom` package. 
:::

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
tissues <- c("Germline", "Neurons", "Muscle", "Hypod.", "Intest.")
df <- map_dfr(tissues, function(tissue) {
    fisher.test(
        REs %over% peaks[['xnd-1']], 
        REs$Annotation == tissue
    ) |> 
        broom::glance() |> 
        mutate(tissue = tissue, TF = 'xnd-1') |> 
        dplyr::select(tissue, TF, estimate, p.value)
})
df
```
:::

- Perform the same operation by iterating over each peak set in the list of imported peaks. 

::: {.callout-tip icon='true'}
- You can use two nested `map_dfr`, iterating first over each TF then over each tissue. 
:::

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
TFs <- names(peaks)
tissues <- c("Germline", "Neurons", "Muscle", "Hypod.", "Intest.")
df <- map_dfr(TFs, function(TF) {
    map_dfr(tissues, function(tissue) {
        fisher.test(
            REs %over% peaks[[TF]], 
            REs$Annotation == tissue
        ) |> 
            broom::glance() |> 
            mutate(tissue = tissue, TF = TF) |> 
            dplyr::select(tissue, TF, estimate, p.value)
    })
})
df
```
:::

- For each TF, filter the tissues in which it is preferentially enriched (odds ratio >= 2, p.value <= 0.05)
- Find TFs enriched over Intestine REs

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
filter(df, estimate >= 2, p.value <= 0.05)
filter(df, estimate >= 2, p.value <= 0.05, tissue == "Intest.")
```
:::

- Check the [STRING DB website](https://string-db.org/cgi/input?input_page_active_form=multiple_identifiers) to assess whether these TFs have been shown to interact together. Comment. 
