---
engine: knitr
execute:
    eval: false
---

# Lab 6: Motif enrichment analysis from ChIP-seq data

::: {.callout-note icon='true'}
## Aims

- Find motifs enriched in a set of ChIP-seq peaks
- Scan genomic loci for de novo motifs
- Compute statistical enrichment of motifs in sets of genomic annotations
:::

::: {.callout-tip icon='true'}
## Datasets

- modENCODE/modERN TF ChIP-seq database, available [here](https://www.encodeproject.org/search/?type=Experiment&control_type!=*&status=released&replicates.library.biosample.donor.organism.scientific_name=Caenorhabditis+elegans&biosample_ontology.term_name=whole+organism&assay_title=TF+ChIP-seq&biosample_ontology.classification=whole%20organisms)
:::

## Find de novo motifs enriched in `xnd-1` ChIP-seq

- Check the [`meme` website](https://meme-suite.org/) to identify which tool is best suited to identify a motif *de novo* in a set of peaks from a ChIP-seq experiment. 
- What do you need to run `xstreme` on a set of peaks? 

### Preparing `meme` input

Extract the genomic sequences for XND-1 peaks using `bedtools getfasta`. 

::: {.callout-answer .icon .callout-note collapse=true}
```{bash filename = 'sh'}
bedtools getfasta -fi Share/genome/ce11.fa -bed Share/modENCODE/xnd-1.bed -name > xnd-1.fa
```
:::

### Running `meme`

- Identify motifs enriched in `xnd-1` peaks with `xstreme` using the following options: 

    - Zero or one occurence per sequence at most for `meme`
    - At most 3 motifs
    - A min motif width of 6 for `meme`
    - A max motif width of 15 for `meme`
    - No `streme` motifs
    - Over multiple processors in parallel

::: {.callout-answer .icon .callout-note collapse=true}
```{bash filename = 'sh'}
xstreme \
    --p xnd-1.fa \
    --oc meme_out/ \
    -minw 12 \
    -maxw 16 \
    --meme-nmotifs 3 \
    --meme-p 12 \
    --meme-mod zoops \
    --streme-nmotifs 0
```
:::

- Check the results. Do they make sense in the light of recent publications? (see [here](https://www.science.org/doi/10.1126/sciadv.abo4082))

## Scan `ce11` regulatory elements for de novo motifs

- Import `ce11` tissue-specific regulatory element annotations, as done in the previous lab.

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(rtracklayer)
library(GenomicRanges)
REs <- readxl::read_xlsx('Share/Celegans/WBcel235_REs.xlsx', skip = 2, col_names = TRUE)
REs <- makeGRangesFromDataFrame(
    REs, seqnames.field = 'chrom_ce11', 
    start.field = 'start_ce11', end.field = 'end_ce11',
    keep.extra.columns = TRUE
)
```
:::

- Import the first `xnd-1` motif found by XSTREME in R using `universalmotif` package.

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(universalmotif)
xnd1_motifs <- read_meme('Share/Celegans/meme_out/xstreme.txt')
names(xnd1_motifs) <- c('motif1', 'motif2', 'motif3')
```
:::

- Read `motifmatchr::matchMotifs()` documentation. 
- What class of objects the motifs should be stored into? Coerce `xnd1-` motifs to `PFMatrixList` class.
- Map `xnd-1` motifs to regulatory elements.
- What is the class of the generated object? Its dimensions? What are the features and what are the columns? 

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(motifmatchr)
library(BiocParallel)
library(TFBSTools)
register(MulticoreParam(workers = 2, progressbar = TRUE))
xnd1_motifs_pfm <- map(xnd1_motifs, convert_motifs, "TFBSTools-PFMatrix") |> do.call(PFMatrixList, args = _)
motif_mappings <- matchMotifs(xnd1_motifs_pfm, REs, genome = BSgenome.Celegans.UCSC.ce11::BSgenome.Celegans.UCSC.ce11)
class(motif_mappings)
dim(motif_mappings)
```
:::

- Compute a fisher test of enrichment for `xnd-1` first motif over Intestine-specific regulatory elements.
- Is `xnd-1` motif enriched in Intestine-specific regulatory elements?

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
has_xnd1 <- motifMatches(motif_mappings)[, 'motif1']
is_intestine <- REs$Annotation == 'Intest.'
tab <- table(is_intestine, has_xnd1)
fisher_test <- fisher.test(tab)
fisher_test
```
:::

- Now do the same analysis for Germline specific regulatory elements.
- Is `xnd-1` motif enriched in Germline-specific regulatory elements?

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
has_xnd1 <- motifMatches(motif_mappings)[, 'motif1']
is_germline <- REs$Annotation == 'Germline'
tab <- table(is_germline, has_xnd1)
fisher_test <- fisher.test(tab)
fisher_test
```
:::

- For each motif, identify in which class of tissue-specific regulatory elements they are enriched. 
- Comment on the results.

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(purrr)
library(broom)
groups <- c("Hypod.", "Muscle", "Intest.", "Neurons",  "Germline", "Soma", "Ubiq.")
pvals <- map_dfr(names(xnd1_motifs_pfm), function(motif) {
    map_dfr(groups, function(group) {
        tab <- table(REs$Annotation == group, motifMatches(motif_mappings)[, motif])
        ftest <- fisher.test(tab) |> 
            tidy() |> 
            mutate(motif = motif, group = group) |> 
            dplyr::select(group, motif, estimate, p.value)
    })
})
filter(pvals, p.value < 0.01, estimate > 2) |> print(n = 40)
```
:::

