---
engine: knitr
execute:
    eval: false
---

# Lab 8: Investigating Hi-C contact maps

::: {.callout-note icon='true'}
## Aims

- Compute the distance-dependent contact probability for a Hi-C contact map
- Annotate chromatin loops and borders 
- Visualize Hi-C contact maps 
- Visualize Hi-C maps aggregated over structural features
:::

## Compute a distance-dependent contact probability from a pairs file

We have a set of Hi-C pairs stored in the `~/Share/pairs/HiC.pairs` file. 

- Use `vroom` to import the file in R 
- Use `plyinteractions` to coerce the data frame into a `GInteractions` object

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(vroom)
library(plyinteractions)
df <- vroom("~/Share/pairs/HiC_filtered.pairs", col_names = c("readid", "seqnames1", "start1", "seqnames2", "start2", "strand1", "strand2", "pair_type"), comment = "#")
gi <- as_ginteractions(df, end1 = start1, end2 = start2)
```
:::

- Compute the `pairdist` distance between each anchor of every interaction in the `GInteractions` object. 
- Tally the number of interactions at each distance, and convert it to a probability by dividing the number of interactions at each distance by the total number of interactions.
- Plot the distance-dependent contact probability as a function of distance, using `HiContacts::plotPs`.

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(tidyverse)
gi <- add_pairdist(gi)
df <- as_tibble(gi) |> 
    group_by(pairdist) |>
    tally() |> 
    mutate(prob = n / sum(n))
plotPs(df, aes(x = pairdist, y = prob))
```
:::

## Annotate chromatin loops and borders

We will use the `chromosight` software in the **terminal** to annotate chromatin loops and borders.

- Check `chromosight` help to understand how to use it.
- Annotate separately `loops` and `borders` from the Hi-C contact map using `chromosight detect` with the `--pattern` option. 

::: {.callout-answer .icon .callout-note collapse=true}
```{sh, filename = 'bash'}
chromosight detect --pattern loops Share/maps/HiC.cool HiC_loops
chromosight detect --pattern borders Share/maps/HiC.cool HiC_borders
```
:::

- Check the aggregated plots generated for loops and borders. 
- Check the content of the `tsv` files generated for loops and borders.

## Visualize Hi-C contact maps

We will use the `HiContacts` package in `R` to visualize Hi-C contact maps.

- Create a pointer to the `Share/maps/HiC.cool` file in R. 
- Check the available chromosomes and resolutions stored in the `cool` file. 
- Import data over the 4th chromosome. 
- Use the `plotContactMap` function to visualize the Hi-C contact map.

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
library(HiContacts)
cf <- CoolFile("Share/maps/HiC.cool")
availableChromosomes(cf)
availableResolutions(cf)
hic <- import(cf, focus = 'IV')
plotMatrix(hic)
```
:::

- Import loops and borders annotated by chromosight in R. 
- Plot again the matrix, over a narrower region, and add the loops and borders as annotations.

::: {.callout-answer .icon .callout-note collapse=true}
```{r filename = 'r'}
loops <- read_tsv("HiC_loops.tsv")

```
:::
