---
engine: knitr
execute:
    eval: false
---

# Lab 9: Multi-omics data integration

::: {.callout-note icon='true'}
## Aims

Cohesin is a ring-like protein complex that plays a crucial role in the organization of chromatin.
it is involved in the formation of chromatin loops and the maintenance of sister chromatid cohesion during cell division. 
Its activity is to extrude the DNA fiber through its ring structure, thereby creating chromatin loops 
and organizing the chromatin into higher-order structures.

In yeast, active gene transcription is thought to contribute to translocation of the cohesin 
complex co-transcriptionally. Thus, cohesin complexes could eventually accumulate preferentially
at convergent sites of active transcription, where they could stabilize chromatin loops.

In this lab, we will analyze a dataset from the yeast *Saccharomyces cerevisiae* to study the
role of cohesin in chromatin organization, in G1 and G2M phases of the cell cycle.

The primary aims of this lab are to: 

1. Identify active and inactive convergent transcription loci
2. Understand whether cohesin is enriched at active convergent transcription loci, compared to inactive ones
3. Assess the distribution of nucleosomes at convergent transcription loci occupied by cohesin
4. Quantify chromatin looping over convergent transcription loci

:::

::: {.callout-tip icon='true'}
## Datasets

- RNAseq @ G1: 
  - rep1
  - rep2
  - rep3

- Hi-C:
  - G1
  - G2M

- MNase: 
  - 5'
  - 10'
  - 15'

- Scc1 ChIP-seq
  - IP
  - Input
:::


## Identify active/inactive convergent transcription loci

- Quantify gene expression using `featureCounts`. 

::: {.callout-answer .icon .callout-note collapse=true}
```{sh, filename = 'bash'}
featureCounts \
    -a Share/genome/R64-1-1.gtf \
    -o RNAseq_counts.txt \
    -t transcript \
    -g gene_id \
    -s 2 \
    -p \
    --countReadPairs \
    -P -B -D 10000 -C \
    -T 2 \
    Share/day5/RNA/RNA_rep*.bam
```
:::

- Import the gene annotations and counts into R 
- Compute average `TPM` values for each gene across the three replicates
- Annotate the genes with their `TPM` values

`TPM` values can be computed using the following function:

```{r, filename = 'R'}
countToTpm <- function(counts, effLen) {
    # effLen is the effective length of the gene, stored in the featureCounts output
    rate <- log(counts) - log(effLen)
    denom <- log(sum(exp(rate)))
    exp(rate - denom + log(1e6))
}
```

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(rtracklayer)
genes <- import('Share/genome/R64-1-1.gtf')
genes <- genes[genes$type == 'transcript']
names(genes) <- genes$gene_id
mcols(genes) <- NULL

library(tidyverse)
cnts <- read_tsv('RNAseq_counts.txt', col_names = TRUE, comment = "#")
genes$efflength <- cnts$Length
cnts <- cnts |> 
    select(-c(2:6)) |> 
    column_to_rownames('Geneid') |> 
    as.matrix()

tpms <- apply(cnts, 2, function(x) countToTpm(x, genes$efflength)) |> rowMeans()
genes$TPM <- tpms
genes$TPM_next_gene <- lead(tpms)
```
:::

- Identify convergent transcription loci (i.e. forward genes followed by reverse genes, which are less than 1kb apart)

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
gene_orientation <- as.vector(strand(genes))
genes$is_convergent <- gene_orientation == '+' & lead(gene_orientation) == '-'

intergenic_width <- lead(as.vector(start(genes))) - as.vector(end(genes))

convergent_genes <- genes[genes$is_convergent & intergenic_width < 1000]
```
:::

- Classify convergent transcription loci as active or inactive based on their `TPM` values (> 100 = active, < 100 = inactive)

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
convergent_genes$active <- case_when(
    convergent_genes$TPM > 100 & convergent_genes$TPM_next_gene > 100 ~ 'active',
    convergent_genes$TPM < 100 & convergent_genes$TPM_next_gene < 100 ~ 'inactive', 
    TRUE ~ 'somewhat_active'
)
export(convergent_genes[convergent_genes$active == 'somewhat_active'] , 'somewhat_active_convergent_genes.bed')
export(convergent_genes[convergent_genes$active == 'active'] , 'active_convergent_genes.bed')
export(convergent_genes[convergent_genes$active == 'inactive'] , 'inactive_convergent_genes.bed')
```
:::

## Annotate cohesin peaks at convergent transcription loci

- Annotate cohesin peaks using `macs2` with the IP and Input files bam files. 

::: {.callout-answer .icon .callout-note collapse=true}
```{sh, filename = 'bash'}
macs2 callpeak \
    -t Share/day5/Scc1/Scc1_IP.bam \
    -c Share/day5/Scc1/Scc1_input.bam \
    -f BAMPE \
    -n Scc1
```
:::

- Import the peaks into R
- Check whether convergent genes are enriched for cohesin peaks, compared to non-convergent genes

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
scc1_peaks <- import('Scc1_peaks.narrowPeak')
seqlevels(scc1_peaks, pruning.mode="coarse")  <- seqlevels(genes)
scc1_peaks$at_active_convergent_loci <- scc1_peaks %over% convergent_genes[convergent_genes$active == 'active']
table(scc1_peaks$at_active_convergent_loci)

table(
    is_convergent = genes$is_convergent, 
    is_bound_by_scc1 = resize(genes, 200, fix = 'end') %over% scc1_peaks
) |> fisher.test()
```
:::

## Plot epigenomics signal at convergent transcription loci

- Group convergent genes by their `TPM` class and plot the average cohesin signal for each class using `tidyCoverage`.
- Interpret the results. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(plyranges)
library(tidyCoverage)
features <- lapply(c('active', 'inactive', 'somewhat_active'), function(x) {
    convergent_genes |> 
        filter(active == x) |> 
        resize(1, fix = 'end') |> 
        resize(4000, fix = 'center') 
})
names(features) <- c('active', 'inactive', 'somewhat_active')

tracks <- list(
    RNA = import('RNA_rep1.bw', as = "Rle"), 
    Scc1 = import('Scc1_IP.bw', as = "Rle")
)

covexp <- CoverageExperiment(tracks, features)
aggcov <- aggregate(covexp)
as_tibble(aggcov) |> 
    ggplot(aes(col = features)) + 
    geom_aggrcoverage() + 
    facet_grid(~ track, scales = 'free')
```
:::

- Import MNase fragments in R from the bam file of each timepoint

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(GenomicAlignments)
mnase_frags <- c(
    readGAlignmentPairs("Share/day5/MNase/MNase_5min.bam"), 
    readGAlignmentPairs("Share/day5/MNase/MNase_10min.bam"),
    readGAlignmentPairs("Share/day5/MNase/MNase_20min.bam")
) |> granges()
mnase_cov <- mnase_frags |> 
    resize(40, fix = 'center') |>
    coverage()
```
:::

- Now only focusing on Scc1 peaks over convergent transcription loci, check the nucleosome distribution using the `MNase` data.

::: {.callout-answer .icon .callout-note collapse=true}
```{r}
features <- scc1_peaks |> 
    filter_by_overlaps(convergent_genes) |> 
    mutate(start = start + peak, end = start) |> 
    resize(2000, fix = 'center') 

tracks <- list(
    RNA = import('RNA_rep1.bw', as = "Rle"), 
    Scc1 = import('Scc1_IP.bw', as = "Rle"), 
    MNase = mnase_cov
)

covexp <- CoverageExperiment(tracks, features)
aggcov <- aggregate(covexp)
as_tibble(aggcov) |> 
    ggplot(aes(col = track)) + 
    geom_aggrcoverage() + 
    facet_grid(~ track, scales = 'free')
```
:::

## Sequence analysis of Scc1 peaks

- Search for motifs in the Scc1 peaks, centered at their summit ± 150 using `STREME`.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(Biostrings)
scc1_summits <- scc1_peaks |> 
    filter_by_overlaps(convergent_genes) |> 
    mutate(start = start + peak, end = start) |> 
    resize(300, fix = 'center') 
genome <- readDNAStringSet("Share/genome/R64-1-1.fa")
seqs <- genome[scc1_summits]
names(seqs) <- as.character(scc1_summits)
writeXStringSet(seqs, 'scc1_summits.fasta')
```

```{sh, filename = 'bash'}
xstreme \
    --order 2 \
    --oc streme_out \
    --meme-nmotifs 0 \
    --fimo-skip \
    --p scc1_summits.fasta
```
:::

- Extract the cut site positions from the MNase data and plot the MNase cut sites pile-up ± 100bp over Scc1 peaks.

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
mnase_cut_5p <- resize(mnase_frags, 1, fix = 'start')
mnase_cut_3p <- resize(mnase_frags, 1, fix = 'end')
mnase_cut <- c(mnase_cut_5p, mnase_cut_3p) |> coverage() 

features <- scc1_peaks |> 
    filter_by_overlaps(convergent_genes) |> 
    mutate(start = start + peak, end = start) |> 
    resize(200, fix = 'center') 

tracks <- mnase_cut

pileup <- CoverageExperiment(tracks, features) |> 
    expand() |> 
    group_by(track, features, coord.scaled) |>
    tally(coverage)

ggplot(pileup, aes(x = coord.scaled, y = n)) + 
    geom_line() 
```
:::

- Comment these results.

## Quantify chromatin looping over Scc1-bound/free convergent transcription loci

- Generate all possible pairs of Scc1 peaks overlapping convergent transcription loci

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(InteractionSet)
library(plyinteractions)
gi_with_scc1 <- GInteractions(
    anchor1 = combn(1:length(scc1_summits), 2)[1, ], 
    anchor2 = combn(1:length(scc1_summits), 2)[2, ], 
    regions = scc1_summits
) |> filter(seqnames1 == seqnames2, abs(start2-end1) >= 15000, abs(start2-end1) <= 30000)
```
:::

- Import G1 and G2/M Hi-C data into R

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(HiContacts)
cfs <- list(
    G1 = CoolFile('Share/day5/HiC/HiC_G1.mcool'), 
    G2M = CoolFile('Share/day5/HiC/HiC_G2M.mcool')
)
```
:::

- Compute the aggregated signal of Hi-C data over the Scc1-bound convergent transcription loci

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
hic <- import(cfs[['G1']], resolution = 1000)
g1_aggr <- aggregate(hic, targets = gi_with_scc1, flankingBins = 9)
hic <- import(cfs[['G2M']], resolution = 1000)
g2m_aggr <- aggregate(hic, targets = gi_with_scc1, flankingBins = 9)
p <- cowplot::plot_grid(
    plotMatrix(g1_aggr, use.scores = 'detrended', scale = 'linear', limits = c(-0.25, 0.25), cmap = bgrColors()) + ggtitle("G1, Scc1 pairs"), 
    plotMatrix(g2m_aggr, use.scores = 'detrended', scale = 'linear', limits = c(-0.25, 0.25), cmap = bgrColors()) + ggtitle("G2M, Scc1 pairs")
)
```
:::

- Comment the results, and propose a unified, multi-step model for how and when chromatin loops are formed in yeast. 
