---
engine: knitr
execute:
    eval: false
---

# Lab 4 - RNA-seq downstream analysis

::: {.callout-note icon='true'}
## Aims

- Estimate transcript abundance from RNA-seq samples (`bam` files)
- Perform differential expression analysis with `DESeq2`
- Identify gene ontology pathways over-represented in genes up-regulated upon heat shock treatment
:::

::: {.callout-tip icon='true'}
## Datasets

RNA-seq data was published in [Nuño-Cabanes et al., Scientific Data 2020](https://www.nature.com/articles/s41597-020-0412-z)

- Control RNA-seq @ 30 C: 

    - `SRR9929263`
    - `SRR9929264`
    - `SRR9929273`
    - `SRR9929282`

- Heat shock RNA-seq @ 39 C, 20 min: `SRR2045248` & `SRR2045249`

    - `SRR9929271`
    - `SRR9929265`
    - `SRR9929280`
    - `SRR9929274`
:::

## Counting mapped fragments over a set of gene annotations 

- Create a tibble containing the following information for the four RNA-seq samples (bam files provided in the shared Google Drive): 

    - Path to local `bam` file
    - The biological sample it corresponds to (e.g. `WT`, `HS`)
    - The biological replicate it corresponds to (e.g. `rep1`, `rep2`, ...)

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(tidyverse)
samples <- tibble(
    bam = list.files('data/mapping', pattern = 'RNAseq.*_R64-1-1.bam$', full.names = TRUE), 
    sample = basename(bam), 
) |> separate(sample, sep = '_', into = c('type', 'sample', 'rep', 'genome'))
```
:::

- Import gene annotations downloaded on Monday for yeast. 
- Filter it out to only keep transcripts

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(rtracklayer)
genes <- import('data/genome/R64-1-1.gtf')
genes <- genes[genes$type == 'transcript']
names(genes) <- genes$gene_name
```
:::

- Check the `summarizeOverlaps()` function from the `GenomicAlignments` package. 
- Run it to compute transcript abundance for all transcripts across the four replicates of the two samples. 
- Specify `BPPARAM` argument to perform RNA counting over multiple CPUs. 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(Rsamtools)
library(GenomicAlignments)
library(SummarizedExperiment)
library(BiocParallel)
RNAseq_counts <- summarizeOverlaps(
    reads = BamFileList(samples$bam), 
    features = genes, 
    singleEnd = FALSE, 
    fragments = TRUE,
    BPPARAM = MulticoreParam(workers = length(samples$bam), progressbar = TRUE)
)
colData(RNAseq_counts) <- DataFrame(samples)
```
:::

## Differential expression analysis

- Read `DESeq` documentation (`?DESeq`). What type of object does it require? How can you create one? 
- Create a `DESeqDataSet` object from the RNA counts. Choose your `design` formula appropriately.
- Run `DESeq` workflow 

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
library(DESeq2)
dds <- DESeqDataSet(RNAseq_counts, design = ~ sample)
dds <- DESeq(dds)
```
:::

- Check the documentation for the `results` function from `DESeq2` package.
- Extract the results of differential expression analysis, for the appropriate `contrast`. 
- Recover genes over-expressed in heat-shock vs. control growth (fold-change >= 2, p-value <= 0.01)

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
res <- results(dds, contrast = c('sample', 'HS', 'WT'))
upreg_genes <- res |> 
    as_tibble(rownames = 'gene') |> 
    filter(log2FoldChange > 1, padj <= 0.01) |> 
    pull(gene)
```
:::

## Gene ontology over-enrichment analysis

- Perform gene ontology enrichment analysis using the `gprofiler2` package. 
- From the results, recover GO terms from the `KEGG` database and extract the most enriched terms
- Comment

::: {.callout-answer .icon .callout-note collapse=true}
```{r, filename = 'R'}
go <- gprofiler2::gost(genes, organism = 'scerevisiae')
go$result |> filter(source == 'KEGG') |> arrange(p_value)
```
:::
